Omics-based Single Step Trait Prediction

### 2017-09-07 paper/manuscript.tex
paper, inbreds as Exp1, shift order of Exp1,Exp2

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp1}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

\Genetics2level{Maize diversity panel}
To address objectives 2 and 3, a maize inbred line data set, subsequently
referred to as \textit{Exp2}, was used~\cite{Yang2014}.
\textit{Exp2} originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

@@

\Genetics2level{Population structure}
Population structure in \textit{Exp2} was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the $l$-th group of inbred lines 
($\mathbf{W}_{l}$) has dimensions $n_{l} \times p$ where $n_{l}$ pertains to 
the number of genotypes in the $l$-th group and $p$ pertains to the number of 
features.
Note that $l$ is only defined for \textit{Exp1} where a distinction between 
the two heterotic groups is necessary whereas $l$ can be ignored for all 
models pertaining to the inbred lines of the maize diversity panel in 
\textit{Exp2}.
To avoid mentioning special cases, we will use the general notation with 
the index $l$ throughout this section, though.
For the diversity panel maize lines, $n_{l}$ corresponds to the number of 
genotypes whereas, in the case of the hybrid data, $n_{l}$ corresponds to the 
number of parent lines in the $l$-th heterotic group.
In generating kernels for each predictor, all features in $\mathbf{W}_{l}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from \textit{Exp1}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp2}, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values).
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from \textit{Exp2}
(Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the superset of genotypes that were
covered by both, genomic and trancriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x = {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was declared
as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either
  \textit{Exp1} or \textit{Exp2}.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar depicts that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed.
  The four bars to the right outline the principle of generating core sets for
  \textit{Exp2}.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the black symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 110 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for two out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.15 as observed for the trait ED\@.
The predictive ability for predictions based on G were always higher for the
full set of inbred lines ($n = 164$) than for the reduced set of inbred lines
($n = 110$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 164 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 110 lines.
In the case of KW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 110 lines than G,
predictive abilities of GT for the full set of 164 lines were also worse than 
those based on G alone.

@@

\Genetics2level{Hybrids}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Core'.
In the `core' set, transcriptomic data were the best single predictor for ADL,
PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the FAT and SUG, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred_core_fraction.pdf}
  \caption{
  Influence of the fraction of genotypes covered by the incomplete predictor
  on predictive abilities.
  Predictive abilities are based on transcriptomic effects derived from a
  single step prediction model involving transcriptomic (T) and genomic (G)
  data, respectively.
  Results are shown for the core set of 110 inbred lines from \textit{Exp2} for
  which both, G and T are available for any genotype and six agronomic traits.
  The fraction of genotypes (\textit{i.e.} `Core Fraction') for which
  T was available is given on the x-axis.
  Dotted lines indicate that $T > G$, dashed lines indicate that $T \approx G$
  and solid lines indicate that $G > T$ when running predictions using single
  predictors on the same data set.
  }
\label{fig:InbredResults}
\end{figure}

@@

\Genetics2level{Superiority of single step models}
\subsection{Exp2}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the core set ($\Delta(n) = 54$) was not large enough to provide
considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to 164 tropical/subtropical lines with little evident
population structure and excluded all other genotypes from our analyses whereas
\citeNP{Guo2016} used genotypes from four different subpopulations, modeled as a
fixed effect.
2) We applied quality checks for the predictor data after generating the subset
of 164 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Exp1}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least for a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
By randomly and repeatedly declaring trancsriptomic information missing from
genotypes in \textit{Exp2} we examined how reliably single step prediction works
depending on the fraction of genotypes covered by both, the complete and the
incomplete predictor.
We observed that about $20--30\%$ of genotypes needed to be covered by both
predictors to match the performance achieved with the complete predictor
individually.
Beyond this fraction of covered genotypes the trend in predictive ability
changes was stable.

>>>>

\section{Materials and Methods}
\Genetics2level{Exp1}
To address objectives 2 and 3, a maize inbred line data set, subsequently
referred to as \textit{Exp1}, was used~\cite{Yang2014}.
\textit{Exp1} originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp2}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

@@

\Genetics2level{Population structure}
Population structure in \textit{Exp1} was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the $l$-th group of inbred lines 
($\mathbf{W}_{l}$) has dimensions $n_{l} \times p$ where $n_{l}$ pertains to 
the number of genotypes in the $l$-th group and $p$ pertains to the number of 
features.
Note that $l$ is only defined for \textit{Exp2} where a distinction between 
the two heterotic groups is necessary whereas $l$ can be ignored for all 
models pertaining to the inbred lines of the maize diversity panel in 
\textit{Exp1}.
To avoid mentioning special cases, we will use the general notation with 
the index $l$ throughout this section, though.
For the diversity panel maize lines, $n_{l}$ corresponds to the number of 
genotypes whereas, in the case of the hybrid data, $n_{l}$ corresponds to the 
number of parent lines in the $l$-th heterotic group.
In generating kernels for each predictor, all features in $\mathbf{W}_{l}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from \textit{Exp2}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp1}, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values).
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from \textit{Exp1}
(Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the superset of genotypes that were
covered by both, genomic and trancriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x = {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was declared
as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either
  \textit{Exp2} or \textit{Exp1}.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar depicts that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed.
  The four bars to the right outline the principle of generating core sets for
  \textit{Exp1}.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the black symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

\section*{Results}
\Genetics2level{Exp1}
At the level of the 110 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for two out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.15 as observed for the trait ED\@.
The predictive ability for predictions based on G were always higher for the
full set of inbred lines ($n = 164$) than for the reduced set of inbred lines
($n = 110$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 164 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 110 lines.
In the case of KW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 110 lines than G,
predictive abilities of GT for the full set of 164 lines were also worse than 
those based on G alone.

@@

\Genetics2level{Exp2}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Core'.
In the `core' set, transcriptomic data were the best single predictor for ADL,
PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the FAT and SUG, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred_core_fraction.pdf}
  \caption{
  Influence of the fraction of genotypes covered by the incomplete predictor
  on predictive abilities.
  Predictive abilities are based on transcriptomic effects derived from a
  single step prediction model involving transcriptomic (T) and genomic (G)
  data, respectively.
  Results are shown for the core set of 110 inbred lines from \textit{Exp1} for
  which both, G and T are available for any genotype and six agronomic traits.
  The fraction of genotypes (\textit{i.e.} `Core Fraction') for which
  T was available is given on the x-axis.
  Dotted lines indicate that $T > G$, dashed lines indicate that $T \approx G$
  and solid lines indicate that $G > T$ when running predictions using single
  predictors on the same data set.
  }
\label{fig:InbredResults}
\end{figure}

@@

\Genetics2level{Superiority of single step models}
\subsection{Exp1}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the core set ($\Delta(n) = 54$) was not large enough to provide
considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to 164 tropical/subtropical lines with little evident
population structure and excluded all other genotypes from our analyses whereas
\citeNP{Guo2016} used genotypes from four different subpopulations, modeled as a
fixed effect.
2) We applied quality checks for the predictor data after generating the subset
of 164 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Exp2}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least for a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
By randomly and repeatedly declaring trancsriptomic information missing from
genotypes in \textit{Exp1} we examined how reliably single step prediction works
depending on the fraction of genotypes covered by both, the complete and the
incomplete predictor.
We observed that about $20--30\%$ of genotypes needed to be covered by both
predictors to match the performance achieved with the complete predictor
individually.
Beyond this fraction of covered genotypes the trend in predictive ability
changes was stable.
